---
title: "Appendix to 'An exploratory evaluation of forecasted changes in sea surface temperature and sea ice in the Domain 1 Marine Protected Area'"
author: "Lucas Krüger, Francisco Santa Cruz, Lorena Rebolledo,César Cárdenas"
date: "`r Sys.Date()`"
output: html_document
---

Evaluate changes of sea surface temperature and sea ice cover within the Domain 1 Marine Protected Area Proposition using CMIP6 data for present observations (1986 to 2005), and future forecasted projections for Medium Term (2041-2060) and Long Term (2081-2100). This evaluation had the goal of starting a preliminary exploratory analysis to: identify climate stable areas that might act as biodiversity refuges; identify future suitable areas for fishing activities and; identify vulnerable areas for climate change effects.



```{r, echo=FALSE}

library(sf)
library(terra)
library(ggplot2)
library(patchwork)
library(dplyr)
library(reshape2)
library(raster)
library(ggsignif)
library(wPerm)
```


load IPCC WGI Interactive atlas data previously downloaded from https://interactive-atlas.ipcc.ch/

Raw Data and Data Description in Iturbide et al. 2021 and 2022, Gutierrez et al. 2021.;  data processing for ensembling in https://github.com/IPCC-WG1/Atlas.git 

Iturbide, M., Fernández, J., Gutiérrez, J.M. et al. Implementation of FAIR principles in the IPCC: the WGI AR6 Atlas repository. Sci Data 9, 629 (2022). https://doi.org/10.1038/s41597-022-01739-y


Iturbide, Maialen, Fernández, Jesús, Gutiérrez, José Manuel, Bedia, Joaquín, Cimadevilla, Ezequiel, Díez-Sierra, Javier, Manzanas, Rodrigo, Casanueva, Ana, Baño-Medina, Jorge, Milovac, Josipa, Herrera, Sixto, Cofiño, Antonio S., San Martín, Daniel, García-Díez, Markel, Hauser, Mathias, Huard, David, & Yelekci, Özge. (2021). Repository supporting the implementation of FAIR principles in the IPCC-WGI Atlas (v2.0-final). Zenodo. https://doi.org/10.5281/zenodo.5171760

Gutiérrez, J.M., R.G. Jones, G.T. Narisma, L.M. Alves, M. Amjad, I.V. Gorodetskaya, M. Grose, N.A.B. Klutse, S.
Krakovska, J. Li, D. Martínez-Castro, L.O. Mearns, S.H. Mernild, T. Ngo-Duc, B. van den Hurk, and J.-H. Yoon,
2021: Atlas. In Climate Change 2021: The Physical Science Basis. Contribution of Working Group I to the Sixth
Assessment Report of the Intergovernmental Panel on Climate Change [Masson-Delmotte, V., P. Zhai, A. Pirani,
S.L. Connors, C. Péan, S. Berger, N.  Caud, Y. Chen, L. Goldfarb, M.I. Gomis, M. Huang, K. Leitzell, E. Lonnoy,
J.B.R. Matthews, T.K. Maycock, T. Waterfield, O. Yelekçi, R. Yu, and B. Zhou (eds.)]. Cambridge University Press,
Cambridge, United Kingdom and New York, NY, USA, pp. 1927–2058, doi:10.1017/9781009157896.021.



```{r echo=TRUE}
#sea ice cover (SIC), seas surface temperature (SST) and surface wind speed (SWS)

sic<-rast("D:/IPCCatlas/Present (1986-2005)/SIC.tiff")
sst<-rast("D:/IPCCatlas/Present (1986-2005)/SST.tiff")


#MT for Medium Term (2041-2060) and Shared Socio-economical Pathways SSP scenarios 1,2,3 & 5

sic_MT1<-rast("D:/IPCCatlas/Medium Term (2041-2060)/SSP1/SIC.tiff")
sst_MT1<-rast("D:/IPCCatlas/Medium Term (2041-2060)/SSP1/SST.tiff")


sic_MT2<-rast("D:/IPCCatlas/Medium Term (2041-2060)/SSP2/SIC.tiff") 
sst_MT2<-rast("D:/IPCCatlas/Medium Term (2041-2060)/SSP2/SST.tiff")


sic_MT3<-rast("D:/IPCCatlas/Medium Term (2041-2060)/SSP3/SIC.tiff") 
sst_MT3<-rast("D:/IPCCatlas/Medium Term (2041-2060)/SSP3/SST.tiff")


sic_MT5<-rast("D:/IPCCatlas/Medium Term (2041-2060)/SSP5/SIC.tiff") 
sst_MT5<-rast("D:/IPCCatlas/Medium Term (2041-2060)/SSP5/SST.tiff")


# LT for  Long Term (2081-2100) and SSP scenarios 1,2,3 & 5

sic_LT1<-rast("D:/IPCCatlas/Long Term (2081-2100)/SSP1/SIC.tiff")
sst_LT1<-rast("D:/IPCCatlas/Long Term (2081-2100)/SSP1/SST.tiff")


sic_LT2<-rast("D:/IPCCatlas/Long Term (2081-2100)/SSP2/SIC.tiff") 
sst_LT2<-rast("D:/IPCCatlas/Long Term (2081-2100)/SSP2/SST.tiff")


sic_LT3<-rast("D:/IPCCatlas/Long Term (2081-2100)/SSP3/SIC.tiff") 
sst_LT3<-rast("D:/IPCCatlas/Long Term (2081-2100)/SSP3/SST.tiff")


sic_LT5<-rast("D:/IPCCatlas/Long Term (2081-2100)/SSP5/SIC.tiff") 
sst_LT5<-rast("D:/IPCCatlas/Long Term (2081-2100)/SSP5/SST.tiff")


names(sic_MT1)<-"SIC_MT1"
names(sst_MT1)<-"SST_MT1"

names(sic_MT2)<-"SIC_MT2"
names(sst_MT2)<-"SST_MT2"

names(sic_MT3)<-"SIC_MT3"
names(sst_MT3)<-"SST_MT3"


names(sic_MT5)<-"SIC_MT5"
names(sst_MT5)<-"SST_MT5"

names(sic_LT1)<-"SIC_LT1"
names(sst_LT1)<-"SST_LT1"

names(sic_LT2)<-"SIC_LT2"
names(sst_LT2)<-"SST_LT2"

names(sic_LT3)<-"SIC_LT3"
names(sst_LT3)<-"SST_LT3"

names(sic_LT5)<-"SIC_LT5"
names(sst_LT5)<-"SST_LT5"


# stack objects 
sic.st<-c(sic,
          sic_MT1,sic_MT2,sic_MT3,sic_MT5,
          
          sic_LT1,sic_LT2,sic_LT3,sic_LT5)

sst.st<-c(sst,
         
         sst_MT1,sst_MT2,sst_MT3,sst_MT5,
         
         sst_LT1, sst_LT2,sst_LT3,sst_LT5)


```

## Domain 1 Marine Protected Area shapefile and points

D1MPA Vertices from https://github.com/drlucaskruger/D1MPA_Vertices.git


```{r, echo=FALSE}

crs<-crs(sic) # coordinate reference system 

d1mpa<-read.csv("D1MPA_points.csv")  # points for sampling SST and SIC

d1sf<-vect(d1mpa,crs=crs,geom=c("Long","Lat")) # convert to spatial vector 

shp<-shapefile("D:/D1MPA/D1MPA from vertices/D1MPA_Vertices/D1MPA_Vertices_2023.shp") # D1MPA shapefile for plotting

shp2<-spTransform(shp,CRSobj = crs) # convert to EPSG 4326 WGS84

shpline<-as(shp2,"SpatialLinesDataFrame") # convert to line

saveRDS(shpline,"D1MPA_contourline.Rds")

```

Extract SST and SIC values to D1MPA points and process dataframe for analysis

```{r,echo=TRUE}

sicv<-terra::extract(sic.st,d1sf,method="simple")
sstv<-terra::extract(sst.st,d1sf,method="simple")

sicdf<-data.frame(sicv,d1mpa[2:4])
sstdf<-data.frame(sstv,d1mpa[2:4])

head(sicdf)

sstN<-subset(sstdf,Sector!="SWAP")
sstS<-subset(sstdf,Sector=="SWAP")
sicN<-subset(sicdf,Sector!="SWAP")
sicS<-subset(sicdf,Sector=="SWAP")


### sea surface temperature 

# north 
ptmN1<-perm.paired.loc(sstN$SST,sstN$SST_MT1,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
ptmN2<-perm.paired.loc(sstN$SST,sstN$SST_MT2,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
ptmN3<-perm.paired.loc(sstN$SST,sstN$SST_MT3,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
ptmN5<-perm.paired.loc(sstN$SST,sstN$SST_MT5,variable="SST change",alternative = "two.sided",R=999,parameter= mean)

ptlN1<-perm.paired.loc(sstN$SST,sstN$SST_LT1,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
ptlN2<-perm.paired.loc(sstN$SST,sstN$SST_LT2,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
ptlN3<-perm.paired.loc(sstN$SST,sstN$SST_LT3,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
ptlN5<-perm.paired.loc(sstN$SST,sstN$SST_LT5,variable="SST change",alternative = "two.sided",R=999,parameter= mean)

# south

ptmS1<-perm.paired.loc(sstS$SST,sstS$SST_MT1,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
ptmS2<-perm.paired.loc(sstS$SST,sstS$SST_MT2,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
ptmS3<-perm.paired.loc(sstS$SST,sstS$SST_MT3,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
ptmS5<-perm.paired.loc(sstS$SST,sstS$SST_MT5,variable="SST change",alternative = "two.sided",R=999,parameter= mean)

ptlS1<-perm.paired.loc(sstS$SST,sstS$SST_LT1,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
ptlS2<-perm.paired.loc(sstS$SST,sstS$SST_LT2,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
ptlS3<-perm.paired.loc(sstS$SST,sstS$SST_LT3,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
ptlS5<-perm.paired.loc(sstS$SST,sstS$SST_LT5,variable="SST change",alternative = "two.sided",R=999,parameter= mean)


### sea ice cover 

# north 
psmN1<-perm.paired.loc(sicN$SIC,sicN$SIC_MT1,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
psmN2<-perm.paired.loc(sicN$SIC,sicN$SIC_MT2,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
psmN3<-perm.paired.loc(sicN$SIC,sicN$SIC_MT3,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
psmN5<-perm.paired.loc(sicN$SIC,sicN$SIC_MT5,variable="SST change",alternative = "two.sided",R=999,parameter= mean)

pslN1<-perm.paired.loc(sicN$SIC,sicN$SIC_LT1,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
pslN2<-perm.paired.loc(sicN$SIC,sicN$SIC_LT2,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
pslN3<-perm.paired.loc(sicN$SIC,sicN$SIC_LT3,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
pslN5<-perm.paired.loc(sicN$SIC,sicN$SIC_LT5,variable="SST change",alternative = "two.sided",R=999,parameter= mean)

# south

psmS1<-perm.paired.loc(sicS$SIC,sicS$SIC_MT1,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
psmS2<-perm.paired.loc(sicS$SIC,sicS$SIC_MT2,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
psmS3<-perm.paired.loc(sicS$SIC,sicS$SIC_MT3,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
psmS5<-perm.paired.loc(sicS$SIC,sicS$SIC_MT5,variable="SST change",alternative = "two.sided",R=999,parameter= mean)

pslS1<-perm.paired.loc(sicS$SIC,sicS$SIC_LT1,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
pslS2<-perm.paired.loc(sicS$SIC,sicS$SIC_LT2,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
pslS3<-perm.paired.loc(sicS$SIC,sicS$SIC_LT3,variable="SST change",alternative = "two.sided",R=999,parameter= mean)
pslS5<-perm.paired.loc(sicS$SIC,sicS$SIC_LT5,variable="SST change",alternative = "two.sided",R=999,parameter= mean)


tchange<-rbind(
data.frame(SSP="SSP1",Sector="N",Term="MT",diff=ptmN1$Observed,p.val=ptmN1$p.value),
data.frame(SSP="SSP2",Sector="N",Term="MT",diff=ptmN2$Observed,p.val=ptmN2$p.value),
data.frame(SSP="SSP3",Sector="N",Term="MT",diff=ptmN3$Observed,p.val=ptmN3$p.value),
data.frame(SSP="SSP5",Sector="N",Term="MT",diff=ptmN5$Observed,p.val=ptmN5$p.value),

data.frame(SSP="SSP1",Sector="N",Term="LT",diff=ptlN1$Observed,p.val=ptlN1$p.value),
data.frame(SSP="SSP2",Sector="N",Term="LT",diff=ptlN2$Observed,p.val=ptlN2$p.value),
data.frame(SSP="SSP3",Sector="N",Term="LT",diff=ptlN3$Observed,p.val=ptlN3$p.value),
data.frame(SSP="SSP5",Sector="N",Term="LT",diff=ptlN5$Observed,p.val=ptlN5$p.value),

data.frame(SSP="SSP1",Sector="S",Term="MT",diff=ptmS1$Observed,p.val=ptmS1$p.value),
data.frame(SSP="SSP2",Sector="S",Term="MT",diff=ptmS2$Observed,p.val=ptmS2$p.value),
data.frame(SSP="SSP3",Sector="S",Term="MT",diff=ptmS3$Observed,p.val=ptmS3$p.value),
data.frame(SSP="SSP5",Sector="S",Term="MT",diff=ptmS5$Observed,p.val=ptmS5$p.value),

data.frame(SSP="SSP1",Sector="S",Term="LT",diff=ptlS1$Observed,p.val=ptlS1$p.value),
data.frame(SSP="SSP2",Sector="S",Term="LT",diff=ptlS2$Observed,p.val=ptlS2$p.value),
data.frame(SSP="SSP3",Sector="S",Term="LT",diff=ptlS3$Observed,p.val=ptlS3$p.value),
data.frame(SSP="SSP5",Sector="S",Term="LT",diff=ptlS5$Observed,p.val=ptlS5$p.value))

summary(tchange$p.val)


schange<-rbind(
data.frame(SSP="SSP1",Sector="N",Term="MT",diff=psmN1$Observed,p.val=psmN1$p.value),
data.frame(SSP="SSP2",Sector="N",Term="MT",diff=psmN2$Observed,p.val=psmN2$p.value),
data.frame(SSP="SSP3",Sector="N",Term="MT",diff=psmN3$Observed,p.val=psmN3$p.value),
data.frame(SSP="SSP5",Sector="N",Term="MT",diff=psmN5$Observed,p.val=psmN5$p.value),

data.frame(SSP="SSP1",Sector="N",Term="LT",diff=pslN1$Observed,p.val=pslN1$p.value),
data.frame(SSP="SSP2",Sector="N",Term="LT",diff=pslN2$Observed,p.val=pslN2$p.value),
data.frame(SSP="SSP3",Sector="N",Term="LT",diff=pslN3$Observed,p.val=pslN3$p.value),
data.frame(SSP="SSP5",Sector="N",Term="LT",diff=pslN5$Observed,p.val=pslN5$p.value),

data.frame(SSP="SSP1",Sector="S",Term="MT",diff=psmS1$Observed,p.val=psmS1$p.value),
data.frame(SSP="SSP2",Sector="S",Term="MT",diff=psmS2$Observed,p.val=psmS2$p.value),
data.frame(SSP="SSP3",Sector="S",Term="MT",diff=psmS3$Observed,p.val=psmS3$p.value),
data.frame(SSP="SSP5",Sector="S",Term="MT",diff=psmS5$Observed,p.val=psmS5$p.value),

data.frame(SSP="SSP1",Sector="S",Term="LT",diff=pslS1$Observed,p.val=pslS1$p.value),
data.frame(SSP="SSP2",Sector="S",Term="LT",diff=pslS2$Observed,p.val=pslS2$p.value),
data.frame(SSP="SSP3",Sector="S",Term="LT",diff=pslS3$Observed,p.val=pslS3$p.value),
data.frame(SSP="SSP5",Sector="S",Term="LT",diff=pslS5$Observed,p.val=pslS5$p.value))

summary(schange$p.val)

tchange$change<-tchange$diff*(-1) # difference is the current - future
schange$change<-schange$diff*(-1) # difference is the current - future


tchange$Term<-factor(tchange$Term,levels=c("MT","LT"))
schange$Term<-factor(schange$Term,levels=c("MT","LT"))

tchange$Year[tchange$Term=="MT"]<-"2041-2060"
tchange$Year[is.na(tchange$Year)]<-"2081-2100"

schange$Year[schange$Term=="MT"]<-"2041-2060"
schange$Year[is.na(schange$Year)]<-"2081-2100"


(ggplot(tchange,aes(Year,change,fill=Sector))+geom_boxplot()+theme_bw()+
  ylab("°C")+
  scale_fill_manual(values=c("red","lightblue"))+
    ggtitle("a. Mean SST change"))+
  
(ggplot(schange,aes(Year,change,fill=Sector))+geom_boxplot()+theme_bw()+
  ylab("%")+
  scale_fill_manual(values=c("red","lightblue"))+
   ggtitle("b. Mean SIC change"))

```



Mapping



```{r,echo=TRUE}


library(reshape2)
msic<-melt(sicdf,id.vars=c("ID","Long","Lat","Sector","SIC"))
msst<-melt(sstdf,id.vars=c("ID","Long","Lat","Sector","SST"))

mdf<-data.frame(msic[2:4],
                Term=substring(msic$variable,first=5,last=6),
                SSP=substring(msic$variable,first=7,last=7),
                SIC=msic$SIC,
                SICproj=msic$value,
                SST=msst$SST,SSTproj=msst$value)


mdf$Sector[mdf$Sector=="SOI" | mdf$Sector=="NWAP"]<-"NSOI" #NorthWestern Antarctic Peninsula and South Orkney Island together


mdf$Term<-factor(mdf$Term,levels=c("MT","LT"))

mdf$sic.ch<-mdf$SICproj-mdf$SIC
mdf$sst.ch<-mdf$SSTproj-mdf$SST

mdf$SSP[mdf$SSP=="1"]<-"SSP 1"
mdf$SSP[mdf$SSP=="2"]<-"SSP 2"
mdf$SSP[mdf$SSP=="3"]<-"SSP 3"
mdf$SSP[mdf$SSP=="5"]<-"SSP 5"

mdf$ss<-paste(mdf$Sector,mdf$SSP)




mdf2<-subset(mdf,SSP!="SSP 1") # subset only scenarios 2,3 and 5


(ggplot()+geom_raster(data=mdf2,aes(x=Long,y=Lat,fill=sst.ch))+
  scale_fill_distiller(palette = "Spectral",direction=-1,name="°C change")+
  geom_path(data=shpline,aes(x = long, y = lat, group = group))+
    coord_quickmap()+theme_bw()+facet_wrap(Term~.)+xlab("Longitude")+ylab("Latitude")+
  ggtitle(label="a. Medium and long Term change in SST"))/



(ggplot()+geom_raster(data=mdf2,aes(x=Long,y=Lat,fill=sic.ch))+
  scale_fill_distiller(palette = "Spectral",direction=+1,name="% change")+
  geom_path(data=shpline,aes(x = long, y = lat, group = group))+
  coord_quickmap()+theme_bw()+facet_wrap(Term~.)+xlab("Longitude")+ylab("Latitude")+
  ggtitle(label="b. Medium and long Term change in SIC"))

```